---
title: "NHANES III (1988-1994): Fixed-Width Files, Cotinine, and Mortality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NHANES III (1988-1994): Fixed-Width Files, Cotinine, and Mortality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

NHANES III (1988-1994) predates the continuous survey. Its public-use data
are not per-cycle XPT files but a few large fixed-width `.dat` files, each
with a SAS program (`.sas`) giving the column positions. `nhanesR` handles the
NHANES III **mortality** file directly (`nhanes_mortality_parse(cycles =
"1988-1994")`); this vignette shows how to read the survey files themselves and
link them.

The example builds an adult analysis file with measured height, weight and
waist, cigarette smoking status checked against serum cotinine, and mortality
follow-up through 2019. NHANES III is worth the extra work for survival
analysis: participants have a median of about 27 years of follow-up.

**None of the code chunks run automatically.** The exam file alone is about
195 MB.

## 1. Download the files

The main files are in `nhanes3/1a/`. Serum cotinine is in the *second*
laboratory file, `nhanes3/2a/lab2.dat`. The main `lab.dat` has a `COP` column,
but it is blank for every record.

```{r download}
raw <- file.path(tempdir(), "nhanes3")
dir.create(raw, showWarnings = FALSE)
options(timeout = 900)                  # default 60 s is too short

base <- "https://wwwn.cdc.gov/nchs/data/nhanes3/"
files <- c("1a/exam.dat",  "1a/exam.sas",
           "1a/adult.dat", "1a/adult.sas",
           "2a/lab2.dat",  "2a/lab2.sas")
for (f in files) {
  dest <- file.path(raw, basename(f))
  if (!file.exists(dest)) download.file(paste0(base, f), dest, mode = "wb")
}
```

Each data file has a codebook alongside it, named `<file>-acc.pdf`
(for example `1a/exam-acc.pdf`). The codebooks list value labels and
missing-value codes.

## 2. Read column positions from the SAS program

Each `.sas` file has an `INPUT` block of lines like `SEQN 1-5` or
`HSSEX 15`. The helper below parses that block and checks it against the
record length (`LRECL`, which counts 2 bytes for the line ending).
`SAScii::parse.SAScii()` produces the same layout.

```{r layout}
sas_layout <- function(sas_file) {
  s <- readLines(sas_file, warn = FALSE)
  lrecl <- as.integer(sub(".*LRECL\\s*=\\s*([0-9]+).*", "\\1",
                          grep("LRECL\\s*=", s, value = TRUE)[1]))
  s <- s[(grep("^\\s*INPUT\\s*$", s)[1] + 1):length(s)]
  s <- s[seq_len(grep("^\\s*;\\s*$", s)[1] - 1)]
  m <- regmatches(s, regexec(
    "^\\s*([A-Z][A-Z0-9_]*)\\s+\\$?\\s*([0-9]+)(-([0-9]+))?\\s*$", s))
  m <- m[lengths(m) == 5]
  L <- data.frame(var   = vapply(m, `[`, "", 2),
                  start = as.integer(vapply(m, `[`, "", 3)))
  end <- vapply(m, `[`, "", 5)
  L$end <- ifelse(nzchar(end), as.integer(end), L$start)
  stopifnot(max(L$end) == lrecl - 2)
  L
}
```

## 3. Read only the columns you need

The exam file has more than 2,000 columns. Reading a handful by position is
fast; reading all of them is not.

```{r reader}
read_nh3 <- function(stem, vars) {
  L <- sas_layout(file.path(raw, paste0(stem, ".sas")))
  stopifnot(all(vars %in% L$var))
  L <- L[match(vars, L$var), ]
  x <- readr::read_fwf(file.path(raw, paste0(stem, ".dat")),
                       readr::fwf_positions(L$start, L$end, L$var),
                       col_types = readr::cols(.default = "d"))
  attr(x, "width") <- setNames(L$end - L$start + 1, L$var)
  x
}

exam  <- read_nh3("exam",  c("SEQN", "HSSEX", "HSAGEIR", "DMARETHN",
                             "BMPHT", "BMPWT", "BMPBMI", "BMPWAIST",
                             "WTPFEX6", "SDPPSU6", "SDPSTRA6"))
adult <- read_nh3("adult", c("SEQN", "HAR1", "HAR3", "HAR16", "HAR24", "HAR27"))
lab2  <- read_nh3("lab2",  c("SEQN", "COP"))
```

Use the 6-year weights (`WTPFEX6`, `SDPPSU6`, `SDPSTRA6`) for the whole
survey. The `*1` and `*2` versions cover phase 1 (1988-1991) or phase 2
(1991-1994) alone.

## 4. Missing-value codes

NHANES III stores "blank but applicable" as all 8s and "don't know" as all
9s, sized to the field: `8`/`9` for a one-character item, `88888` for a
five-character measurement. Left in place they look like real values: an
88888 cotinine code makes the mean serum cotinine about 10,000 ng/mL.

```{r recode}
recode_codes <- function(x, vars) {
  w <- attr(x, "width")
  for (v in vars) {
    codes <- as.numeric(c(strrep("8", w[[v]]), strrep("9", w[[v]])))
    x[[v]][x[[v]] %in% codes] <- NA
  }
  x
}
exam  <- recode_codes(exam,  c("BMPHT", "BMPWT", "BMPBMI", "BMPWAIST"))
adult <- recode_codes(adult, c("HAR1", "HAR3", "HAR16", "HAR24", "HAR27"))
lab2  <- recode_codes(lab2,  "COP")
```

Check each variable's codebook entry before applying this rule to it. Age
(`HSAGEIR`) is top-coded at 90 and has no such codes.

Missing serum cotinine is mostly missing *labs*, not a missing assay: children
under 4 were not eligible, and most people coded 88888 have no serum results
at all.

## 5. Smoking status, checked against cotinine

`HAR1` asks whether the person has smoked 100 or more cigarettes; only those
who say yes are asked `HAR3` (smoke now). Other tobacco items are asked only of
people who ever used that product, so a skipped item means "no".

```{r smoking}
d <- merge(exam, adult, by = "SEQN")            # adults 17+ with an exam
d <- merge(d, lab2, by = "SEQN", all.x = TRUE)

d$cig_status <- with(d, ifelse(HAR1 == 2, "Never",
                        ifelse(HAR1 == 1 & HAR3 == 2, "Former",
                        ifelse(HAR1 == 1 & HAR3 == 1, "Current", NA))))
d$other_tob  <- with(d, HAR16 %in% 1 | HAR24 %in% 1 | HAR27 %in% 1)

## serum cotinine > 10 ng/mL is the usual cutoff for active tobacco use
aggregate(COP ~ cig_status, data = d,
          FUN = function(v) round(c(median = median(v), pct_over10 = 100 * mean(v > 10)), 2))
```

About 6% of self-reported never-smokers have cotinine above 10 ng/mL, and
about 13% of former smokers. A stricter never-smoker definition:

```{r never}
d$never_conf <- with(d, cig_status %in% "Never" & !other_tob &
                        (is.na(COP) | COP <= 10))
```

## 6. Link mortality

`nhanesR` downloads and parses the 2019 public-use Linked Mortality File for
NHANES III like any other cycle.

```{r mortality}
library(nhanesR)
mort <- nhanes_mortality_parse(cycles = "1988-1994")[["1988-1994"]]

d <- merge(d, mort[, c("SEQN", "ELIGSTAT", "MORTSTAT", "UCOD_LEADING",
                       "PERMTH_EXM", "PERMTH_INT")],
           by = "SEQN", all.x = TRUE)
table(d$ELIGSTAT, d$MORTSTAT, useNA = "ifany")

d <- d[d$ELIGSTAT %in% 1 & !is.na(d$PERMTH_EXM), ]
d$time  <- d$PERMTH_EXM                  # months from the exam
d$event <- d$MORTSTAT                    # 1 = died by 2019-12-31
```

`ELIGSTAT` is 2 for people under 18 at the survey (not linked) and 3 when
there was too little identifying information. About 450 linkable adults have
no `PERMTH_EXM`: they were examined at home rather than in the mobile exam
centre, and have a zero MEC exam weight. `PERMTH_INT` gives their follow-up
from the interview if you need them.

## 7. Survey-weighted Cox model

```{r cox}
library(survey)
library(survival)
d$Ht_m <- d$BMPHT / 100
des <- svydesign(ids = ~SDPPSU6, strata = ~SDPSTRA6, weights = ~WTPFEX6,
                 nest = TRUE, data = d[d$WTPFEX6 > 0, ])
fit <- svycoxph(Surv(time, event) ~ HSAGEIR + factor(HSSEX) + BMPBMI + Ht_m,
                design = subset(des, never_conf & !is.na(BMPBMI)))
summary(fit)
```

## Notes

- Public-use mortality files add small perturbations to some follow-up times
  and causes of death for confidentiality. NCHS reports the effect on
  relative-risk estimates is negligible.
- `UCOD_LEADING` uses the same 10-category leading-cause recode as the
  continuous-NHANES files; see `nhanes_ucod_labels()`.
- NHANES III also has youth (`1a/youth.dat`) and first-laboratory
  (`1a/lab.dat`) files; the same reader works for them.
