---
title: "NHANES Mortality Linkage: A Complete Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NHANES Mortality Linkage: A Complete Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Overview

This vignette walks through the complete `nhanesR` workflow:

1. Browse available files with `nhanes_manifest()`
2. Download and cache component XPT files with `nhanes_download()`
3. Stack data across cycles with `nhanes_stack()`
4. Merge components by SEQN with `nhanes_merge()`
5. Link mortality follow-up with `nhanes_mortality_link()`
6. Prepare a survival dataset with `nhanes_survival_prep()`
7. Run a survey-weighted Cox model

All download steps hit the CDC servers and are marked `eval=FALSE` — run them
interactively. Results are cached locally so subsequent calls are fast.

```{r load}
library(nhanesR)
library(survival)
library(survey)
```

---

## 1. What cycles and files are available?

```{r cycles}
# All continuous NHANES cycles
nhanes_cycles()

# Include NHANES III
nhanes_cycles(include_iii = TRUE)
```

`nhanes_manifest()` scrapes the CDC data catalog and returns a table of
available files for a given cycle and component:

```{r manifest}
# What laboratory files exist for 2015-2016?
nhanes_manifest("2015-2016", "Laboratory")

# Demographics for 2013-2014
nhanes_manifest("2013-2014", "Demographics")
```

---

## 2. Download data

We will examine the association between serum total cholesterol and
cardiovascular mortality across four NHANES cycles (2011-2018).

The relevant files are:

| Component | File code | Contents |
|-----------|-----------|----------|
| Demographics | `DEMO` | Age, sex, race, survey weights, design variables |
| Laboratory | `TCHOL` | Serum total cholesterol |

```{r download}
cycles <- c("2011-2012", "2013-2014", "2015-2016", "2017-2018")

# nhanes_download() returns a list when multiple cycles are requested.
# Each element is a data frame with a 'cycle' column added automatically.
demo_list  <- nhanes_download("DEMO",  cycles)
tchol_list <- nhanes_download("TCHOL", cycles)
```

Downloaded files are parsed from SAS transport (XPT) format via `haven` and
cached as RDS files locally. Re-running the above is instant on subsequent
calls.

---

## 3. Stack across cycles

`nhanes_stack()` row-binds the per-cycle data frames. Columns present in some
but not all cycles are filled with `NA` by default (`fill = TRUE`), preserving
all variables across cycles where variable names changed.

```{r stack}
demo  <- nhanes_stack(demo_list)
tchol <- nhanes_stack(tchol_list)

nrow(demo)   # ~40,000 participants across four cycles
nrow(tchol)
```

---

## 4. Merge components

`nhanes_merge()` joins on `SEQN` (respondent sequence number). For pooled
multi-cycle data, always include `"cycle"` in the `by` argument to prevent
cross-cycle SEQN collisions.

```{r merge}
analytic <- nhanes_merge(demo, tchol, by = c("SEQN", "cycle"))

# Survey design variables should now be present
c("SDMVPSU", "SDMVSTRA", "WTMEC2YR") %in% names(analytic)

# Cholesterol variable
summary(analytic$LBXTC)   # mg/dL; NA where not examined
```

---

## 5. Link mortality

`nhanes_mortality_link()` downloads the CDC Public-Use Linked Mortality Files
(LMF) for the relevant cycles, parses the fixed-width `.dat` files, and
left-joins onto your analytic dataset.

Mortality follow-up runs through **December 31, 2019** for all public-use
files (released April 2022).

```{r mortality}
analytic_mort <- nhanes_mortality_link(analytic)

# Key variables added:
# ELIGSTAT  - eligibility for follow-up (1 = eligible)
# MORTSTAT  - vital status (0 = assumed alive, 1 = deceased)
# UCOD_LEADING - underlying cause of death (ICD-10 recode)
# PERMTH_EXM   - months of follow-up from examination date

table(analytic_mort$MORTSTAT, useNA = "always")
```

---

## 6. Prepare for survival analysis

`nhanes_survival_prep()` handles the steps that are easy to get wrong:

- Removes ineligible participants (`ELIGSTAT != 1`) with a warning and count
- Creates `time` (from `PERMTH_EXM` or `PERMTH_INT`) and `event` columns
- Warns about asymmetric follow-up across cycles (later cycles have shorter
  maximum follow-up since all censor at the same date)
- Optionally creates a cause-specific `event_cause` indicator

```{r survprep}
# All-cause mortality, follow-up from examination date
surv_data <- nhanes_survival_prep(
  analytic_mort,
  origin     = "exam",
  time_unit  = "years",
  weight_var = "WTMEC2YR"
)

# How does follow-up look per cycle?
nhanes_followup_summary(surv_data)
```

For a cardiovascular-specific analysis, supply a cause code. Use
`nhanes_ucod_labels()` to look up codes:

```{r ucod}
nhanes_ucod_labels()
```

```{r survprep-cvd}
# Cardiovascular mortality (code "001": Diseases of heart)
surv_cvd <- nhanes_survival_prep(
  analytic_mort,
  origin     = "exam",
  time_unit  = "years",
  cause      = "001",
  weight_var = "WTMEC2YR"
)

# event_cause = 1 only when UCOD_LEADING == "001" AND MORTSTAT == 1
table(surv_cvd$event, surv_cvd$event_cause)
```

---

## 7. Survey-weighted Cox model

NHANES uses a complex multi-stage probability sample. Ignoring the survey
design produces incorrect standard errors. The `survey` package handles this
via `svydesign()` + `svycoxph()`.

**Weight adjustment for pooled cycles**: when combining more than two years of
data, divide the 2-year MEC weight by the number of cycles pooled. Here we
pool four cycles (8 years), so divide `WTMEC2YR` by 4.

```{r survey-design}
surv_data$wt_pooled <- surv_data$survey_weight / 4

design <- svydesign(
  id      = ~SDMVPSU,
  strata  = ~SDMVSTRA,
  weights = ~wt_pooled,
  nest    = TRUE,
  data    = surv_data
)
```

Fit a Cox proportional hazards model for total cholesterol and all-cause
mortality, adjusting for age and sex:

```{r cox}
# Standardise cholesterol to per-SD units for interpretability
surv_data$tchol_sd <- scale(surv_data$LBXTC)[, 1]

# Re-create design with standardised variable
design <- svydesign(
  id      = ~SDMVPSU,
  strata  = ~SDMVSTRA,
  weights = ~wt_pooled,
  nest    = TRUE,
  data    = surv_data
)

fit <- svycoxph(
  Surv(time, event) ~ tchol_sd + RIDAGEYR + RIAGENDR,
  design = design
)

summary(fit)
exp(coef(fit))          # hazard ratios
exp(confint(fit))       # 95% confidence intervals
```

---

## Notes on the public-use LMF

- Follow-up is through **December 31, 2019** for all cycles. Participants from
  later cycles (e.g. 2017-2018) have substantially shorter maximum follow-up
  (~1-2 years) compared to 1999-2000 participants (~20 years). This asymmetry
  must be accounted for in pooled analyses.

- `PERMTH_EXM`, `PERMTH_INT`, and `UCOD_LEADING` contain **synthetic
  (perturbed) values** for select records, introduced by CDC to reduce
  re-identification risk. `MORTSTAT` and `ELIGSTAT` are not perturbed.

- The 2022-linked restricted-use files extend follow-up to December 31, 2022
  and are available only through NCHS Research Data Centers (RDC).

- See the CDC documentation:
  <https://www.cdc.gov/nchs/data-linkage/mortality-public.htm>
