---
title: "NHANES Mortality Linkage: A Complete Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NHANES Mortality Linkage: A Complete Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Overview

This vignette walks through the complete `nhanesR` workflow using a concrete
example: the association between serum total cholesterol and cardiovascular
mortality across nine NHANES cycles (1999–2016).

**None of the code chunks run automatically** — copy and paste each one into
your R console and run it interactively. Downloaded files are cached locally,
so re-running any step is fast after the first time.

The workflow has seven steps:

1. Browse available files with `nhanes_manifest()`
2. Download and cache data files with `nhanes_download()`
3. Stack data across cycles with `nhanes_stack()`
4. Merge components by participant with `nhanes_merge()`
5. Link mortality follow-up with `nhanes_mortality_link()`
6. Prepare a survival dataset with `nhanes_survival_prep()`
7. Run a survey-weighted Cox proportional hazards model

```{r load}
library(nhanesR)
library(survival)
library(survey)
```

---

## Background: NHANES structure

NHANES (National Health and Nutrition Examination Survey) is conducted in
two-year **cycles** (e.g. 1999–2000, 2001–2002, ..., 2017–2018). Within each
cycle, data are organized into five **components**:

| Component | What it contains |
|-----------|-----------------|
| Demographics | Age, sex, race/ethnicity, income, survey weights and design variables |
| Laboratory | Blood and urine measurements |
| Examination | Physical exam, blood pressure, anthropometry |
| Questionnaire | Self-reported health history, medications, behaviors |
| Dietary | 24-hour dietary recall interviews |

Each participant has a unique identifier, **SEQN**, that is used to link files
within a cycle. SEQNs are not reused across cycles, so when pooling multiple
cycles it is essential to keep track of which cycle each SEQN came from.

The **continuous NHANES** (1999–present) uses a consistent file structure and
naming convention. **NHANES III** (1988–1994) predates this and uses a
different structure; `nhanesR` supports it but it requires extra care.

---

## 1. What cycles and files are available?

```{r cycles}
# All continuous NHANES cycles known to nhanesR
nhanes_cycles()

# Add NHANES III (different structure — see ?nhanes_cycles)
nhanes_cycles(include_iii = TRUE)

# Just the cycle labels for the first 9 continuous cycles
nhanes_cycles()[1:9, "cycle"]
```

To see what files are available for a specific cycle and component, use
`nhanes_manifest()`. It queries the CDC data catalog and returns a table of
file codes, descriptions, and download URLs.

```{r manifest}
# What laboratory files exist for 2015-2016?
nhanes_manifest("2015-2016", "Laboratory")

# Demographics files for 2013-2014
nhanes_manifest("2013-2014", "Demographics")
```

The `file_name` column (e.g. `TCHOL_I`, `DEMO_H`) gives the file code you
pass to `nhanes_download()`. The letter suffix (`_I`, `_H`, etc.) identifies
the cycle; `nhanesR` handles this automatically — you only need the stem
(e.g. `"TCHOL"`, `"DEMO"`).

---

## 2. Download data

We need two files per cycle:

| Component | File code | Key variables |
|-----------|-----------|---------------|
| Demographics | `DEMO` | Age (`RIDAGEYR`), sex (`RIAGENDR`), survey weights (`WTMEC2YR`), design variables (`SDMVPSU`, `SDMVSTRA`) |
| Laboratory | `TCHOL` | Serum total cholesterol (`LBXTC`, in mg/dL) |

`RIAGENDR` is coded 1 = male, 2 = female. `SDMVPSU` and `SDMVSTRA` are the
pseudo-PSU and pseudo-stratum variables required for variance estimation under
NHANES's complex multi-stage sampling design — they must come from the
Demographics file and be included in any survey-weighted analysis.

```{r download}
cycles <- nhanes_cycles()[1:9, "cycle"]   # 1999-2016

# When multiple cycles are requested, nhanes_download() returns a named list.
# Each element is a data frame; a 'cycle' column is added automatically.
demo_list  <- nhanes_download("DEMO",  cycles)
tchol_list <- nhanes_download("TCHOL", cycles)
```

Files are downloaded from the CDC in SAS transport (XPT) format, parsed into
R data frames, and saved to a local cache. If you run these lines again, the
cached versions are loaded instead of re-downloading.

To see where the cache lives:

```{r cache}
nhanes_cache_dir()
```

---

## 3. Stack across cycles

`nhanes_stack()` row-binds the per-cycle data frames into a single data frame.
If a variable was renamed or added between cycles, columns present in some
cycles but not others are filled with `NA` rather than dropped.

```{r stack}
demo  <- nhanes_stack(demo_list)
tchol <- nhanes_stack(tchol_list)

nrow(demo)    # approximately 90,000 participants across nine cycles
nrow(tchol)   # smaller — not all participants attended the lab exam
```

---

## 4. Merge components

`nhanes_merge()` joins data frames on SEQN. Because SEQNs are unique only
within a cycle, always include `"cycle"` in the join key when working with
pooled multi-cycle data. An inner join (the default) retains only participants
present in both files — those who completed both the interview and the lab
examination.

```{r merge}
analytic <- nhanes_merge(demo, tchol, by = c("SEQN", "cycle"))

# Confirm that the survey design variables came through from DEMO
c("SDMVPSU", "SDMVSTRA", "WTMEC2YR") %in% names(analytic)

# Check the cholesterol distribution
summary(analytic$LBXTC)   # mg/dL; few NAs expected after inner join
```

---

## 5. Link mortality

`nhanes_mortality_link()` downloads the CDC National Center for Health
Statistics (NCHS) Public-Use Linked Mortality Files (LMF) for the relevant
cycles, and left-joins them onto your analytic dataset by SEQN.

Mortality follow-up in these files runs through **December 31, 2019** for all
cycles. The files were released in April 2022 and will not be updated — more
recent follow-up (to December 31, 2022) is available only through the NCHS
Research Data Center (RDC) as a restricted-use file.

```{r mortality}
analytic_mort <- nhanes_mortality_link(analytic)

# Key variables added by the LMF:
#
#   ELIGSTAT     - Eligibility for mortality follow-up:
#                    1 = eligible
#                    2 = under 18 at time of survey (not followed)
#                    3 = insufficient identifying data for linkage
#
#   MORTSTAT     - Final vital status:
#                    0 = assumed alive as of December 31, 2019
#                    1 = assumed deceased
#
#   UCOD_LEADING - Underlying cause of death (11-category ICD-10 recode);
#                    NA if alive or ineligible
#
#   PERMTH_EXM   - Months of follow-up from examination date to death or
#                    December 31, 2019, whichever came first
#
#   PERMTH_INT   - Same, but from interview date

table(analytic_mort$MORTSTAT, useNA = "always")
```

---

## 6. Prepare for survival analysis

`nhanes_survival_prep()` handles several steps that are easy to get wrong:

- **Removes ineligible participants** (`ELIGSTAT != 1`) — those under 18 and
  those with insufficient identifying data cannot be used in survival analyses
  with the public-use LMF. The number dropped is reported.
- **Creates `time` and `event` columns** ready for `survival::Surv()`.
- **Warns about asymmetric follow-up** — because all cycles censor at the same
  date (December 31, 2019), participants enrolled in later cycles have much
  shorter maximum follow-up than those enrolled in earlier cycles. This must be
  kept in mind when interpreting pooled analyses.

**Choosing `origin`**: follow-up time can be measured from the **interview
date** (`origin = "interview"`, using `PERMTH_INT`) or the **examination date**
(`origin = "exam"`, using `PERMTH_EXM`). When your analysis includes laboratory
or physical examination measurements — as this one does — use `origin = "exam"`,
since those measurements were obtained at the examination visit.

```{r survprep}
surv_data <- nhanes_survival_prep(
  analytic_mort,
  origin     = "exam",
  time_unit  = "years",
  weight_var = "WTMEC2YR"
)

# Summarise follow-up time and event rates by cycle.
# Note the shorter maximum follow-up for later cycles.
nhanes_followup_summary(surv_data)
```

You should see something like:

```
      cycle     n n_events event_rate_pct median_followup max_followup time_unit
1 2011-2012  8xxx      xxx           x.xx            x.x         ~8.x     years
2 2013-2014  8xxx      xxx           x.xx            x.x         ~6.x     years
3 2015-2016  8xxx      xxx           x.xx            x.x         ~4.x     years
4 2017-2018  8xxx      xxx           x.xx            x.x         ~2.x     years
```

The shrinking maximum follow-up across cycles is expected and is a consequence
of the fixed December 31, 2019 censor date.

For a **cause-specific** analysis, use `nhanes_ucod_labels()` to look up the
available cause-of-death codes, then pass the code to `nhanes_survival_prep()`:

```{r ucod}
# 11-category ICD-10 recode used in the public-use LMF
nhanes_ucod_labels()
```

```{r survprep-cvd}
# Cardiovascular mortality only (code "001": Diseases of heart)
surv_cvd <- nhanes_survival_prep(
  analytic_mort,
  origin     = "exam",
  time_unit  = "years",
  cause      = "001",
  weight_var = "WTMEC2YR"
)

# event       = 1 for all-cause death
# event_cause = 1 only for cardiovascular death
# All event_cause == 1 rows should also have event == 1
table(event = surv_cvd$event, cvd_death = surv_cvd$event_cause)
```

---

## 7. Survey-weighted Cox model

NHANES uses a complex multi-stage probability sample. Standard errors computed
without accounting for the sampling design will be wrong — typically
anti-conservative. The `survey` package corrects for this.

**Weight adjustment for pooled cycles**: the MEC exam weight `WTMEC2YR` is
designed for a single two-year cycle. When pooling multiple cycles, the NHANES
analytic guidelines specify dividing the 2-year weight by the number of cycles
combined. Here we pool nine cycles (18 years), so divide by 9.

`nest = TRUE` tells `svydesign()` that PSU identifiers (SDMVPSU) are nested
within strata and may repeat across strata — this is the correct specification
for NHANES.

```{r survey-design}
surv_data$wt_pooled <- surv_data$survey_weight / 9

# Standardise cholesterol to per-SD units so the hazard ratio is interpretable
# as the risk associated with a one-standard-deviation higher cholesterol level
surv_data$tchol_sd <- scale(surv_data$LBXTC)[, 1]

design <- svydesign(
  id      = ~SDMVPSU,
  strata  = ~SDMVSTRA,
  weights = ~wt_pooled,
  nest    = TRUE,
  data    = surv_data
)
```

Fit a Cox model for all-cause mortality, adjusting for age and sex:

```{r cox}
fit <- svycoxph(
  Surv(time, event) ~ tchol_sd + RIDAGEYR + RIAGENDR,
  design = design
)

summary(fit)
exp(coef(fit))     # hazard ratios
exp(confint(fit))  # 95% confidence intervals
```

The hazard ratio for `tchol_sd` estimates the mortality risk associated with a
one-SD higher total cholesterol, adjusted for age and sex, in the U.S.
non-institutionalized civilian population (the population NHANES represents).

`RIAGENDR` is coded 1 = male, 2 = female, so its hazard ratio compares females
to males.

---

## Notes on the public-use LMF

- **Asymmetric follow-up**: all public-use LMF files censor at December 31,
  2019 regardless of cycle. Participants from 2017–2018 have at most ~2 years
  of follow-up; those from 1999–2000 have up to ~20 years. This must be
  accounted for — or at minimum acknowledged — in any pooled analysis.

- **Data perturbation**: `PERMTH_EXM`, `PERMTH_INT`, and `UCOD_LEADING`
  contain synthetic (perturbed) values for select records, introduced by CDC
  to reduce re-identification risk. `MORTSTAT` and `ELIGSTAT` are not
  perturbed. Cause-specific analyses using `UCOD_LEADING` should be
  interpreted with this in mind.

- **Restricted-use files**: the 2022-linked files extend follow-up to
  December 31, 2022 and include additional cause-of-death detail. They require
  an approved project and on-site or virtual RDC access.

- **Further reading**:
  - CDC mortality linkage: <https://www.cdc.gov/nchs/data-linkage/mortality-public.htm>
  - NHANES analytic guidelines: <https://wwwn.cdc.gov/nchs/nhanes/analyticguidelines.aspx>
  - `survey` package documentation: <https://r-survey.r-forge.r-project.org/survey/>
