---
title: "Comparing Survival Frameworks: Cox, Piecewise Exponential, and mgcv-Poisson"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Survival Frameworks: Cox, Piecewise Exponential, and mgcv-Poisson}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE
)
```

> **Note on validation.** This vignette is fully self-contained — it
> simulates its own data and needs no network access or external files —
> and every chunk below has been executed end to end (R 4.3.3, `survival`
> 3.5-8, `mgcv` 1.9-1, knitting via `rmarkdown`/`knitr`/pandoc), not just
> checked by eye. `eval = TRUE` is the default. Numbers quoted in the prose
> (coefficient estimates, the convergence table, the weighted-design
> results) are real output from that run and will vary slightly on a
> re-knit — different R/package versions, and the RNG stream, mean exact
> digits will not reproduce bit-for-bit even with the same `set.seed()`
> call across machines — but the qualitative pattern each paragraph
> describes should reproduce. One earlier design choice in the "Design
> weights" section (an omitted-covariate construction for the "informative
> weighting" regime) did **not** hold up under this validation pass — see
> the note in that section — and was replaced with a cleaner design after
> the discrepancy was traced to its root cause.

## Motivation

Three common frameworks for estimating a covariate-adjusted hazard produce
numerically related, and in one limiting case *identical*, answers:

1. **Cox proportional hazards** (`survival::coxph()`) — semiparametric,
   partial likelihood, baseline hazard left completely unspecified.
2. **Piecewise exponential model (PEM)** — the time axis is cut into
   intervals, the hazard is assumed constant within each interval, and the
   model is fit as a Poisson GLM on person-interval records
   (`stats::glm(family = poisson)`).
3. **mgcv-Poisson / PAMM** (piecewise-exponential *additive* model,
   Bender, Groll & Scheipl 2018) — the same person-interval data as (2), but
   the interval-specific log-hazards are replaced by a penalized smooth
   function of time, fit with `mgcv::gam(family = poisson)`.

These are usually presented as three unrelated modeling choices. They are
not: (2) is a discretized, computationally cheap approximation to (1), and
(3) is what you get when you stop forcing the discretization to be a step
function and let a penalized spline choose its own effective smoothness.
The clearest way to see the relationship is to look at what each framework
contributes to its likelihood **from the risk set present at a single event
time** — the atomic unit every one of these models is built from.

This document derives that contribution for all three frameworks side by
side, shows the algebraic equivalence between (1) and (2) in the limit of
fine time-splitting, and gives a runnable simulation comparing all three
numerically. It then extends the same risk-set-contribution machinery to
survey design weights, and answers the question this comparison project
was originally motivated by: design weights are well known to matter for
population-level means and proportions — why do
`vignettes/survey-weighted-survival.Rmd`'s empirical results show their
effect on Cox regression coefficients to be so much smaller, and so
predictor-specific? See the "Design weights" section below.

### Key references

- Whitehead, J. (1980). Fitting Cox's regression model to survival data
  using GLIM. *Applied Statistics*, 29(3), 268–275.
- Laird, N. & Olivier, D. (1981). Covariance analysis of censored survival
  data using log-linear analysis techniques. *JASA*, 76(374), 231–240.
- Holford, T.R. (1980). The analysis of rates and of survivorship using
  log-linear models. *Biometrics*, 36(2), 299–305.
- Bender, A., Groll, A. & Scheipl, F. (2018). A generalized additive model
  approach to time-to-event analysis. *Statistical Modelling*, 18(3-4),
  299–321. [PAMMs; see also the `pammtools` package]
- Wood, S.N. (2017). *Generalized Additive Models: An Introduction with R*,
  2nd ed. CRC Press. [`mgcv`]
- Therneau, T.M. & Grambsch, P.M. (2000). *Modeling Survival Data: Extending
  the Cox Model*. Springer.
- Bender, R., Augustin, T. & Blettner, M. (2005). Generating survival times
  to simulate Cox proportional hazards models. *Statistics in Medicine*,
  24(11), 1713–1723. [the simulation scheme used below]
- Binder, D.A. (1992). Fitting Cox's proportional hazards models from
  survey data. *Biometrika*, 79(1), 139–147. [the weighted risk-set
  contribution derived in "Design weights" below; see also
  `vignettes/survey-weighted-survival.Rmd`]
- Gail, M.H., Wieand, S. & Piantadosi, S. (1984). Biased estimates of
  treatment effect in randomized experiments with nonlinear regressions
  and omitted covariates. *Biometrika*, 71(3), 431–444. [omitted-covariate
  attenuation and dynamic risk-set selection; the mechanism behind the
  discarded construction noted in "Design weights" below]

---

## Notation

For subject $j$ with covariate vector $x_j$ and linear predictor
$\eta_j = x_j^\top \beta$, let $R(t)$ denote the risk set at time $t$ — the
set of subjects still under observation (not yet failed or censored) just
before $t$. An event occurs at time $t_i$ for subject $i^\*(t_i) \in R(t_i)$.

---

## Cox: the partial-likelihood contribution

Ignoring ties, the Cox partial likelihood contribution at event time
$t_i$ is the now-familiar risk-set ratio:

$$
L_i(\beta) = \frac{\exp(\eta_{i^\*})}{\sum_{j \in R(t_i)} \exp(\eta_j)},
\qquad
\ell_i(\beta) = \eta_{i^\*} - \log \sum_{j \in R(t_i)} \exp(\eta_j)
$$

The baseline hazard $\lambda_0(t)$ never appears: it cancels out of the
ratio entirely. This is the sense in which Cox regression is
semiparametric — $\lambda_0(t)$ is an infinite-dimensional nuisance
parameter that is *profiled out analytically*, not estimated. Ties are
handled by the Breslow or Efron approximation to the exact discrete-risk-set
likelihood; `coxph()` uses Efron by default.

---

## Piecewise exponential: the Poisson-trick contribution

Partition the time axis into intervals with cutpoints
$0 = c_0 < c_1 < \dots < c_K$. Every subject contributes one
person-interval record for each interval they were at risk in. For
subject $j$ in interval $k$: exposure time $o_{jk}$ (time spent at risk in
that interval), event indicator $d_{jk} \in \{0, 1\}$ (1 only if the event
happened in that interval), and assumed constant hazard within the
interval, $\lambda_k(x_j) = \lambda_{0k}\exp(\eta_j)$. Then
$d_{jk} \sim \text{Poisson}(\mu_{jk})$ with
$\mu_{jk} = o_{jk}\lambda_{0k}\exp(\eta_j) = \exp(\log o_{jk} + \alpha_k + \eta_j)$,
i.e. a Poisson GLM with an offset of $\log o_{jk}$ and a free intercept
$\alpha_k = \log \lambda_{0k}$ per interval.

For an interval $k$ containing exactly one event, at subject $i^\*$, the
Poisson log-likelihood contributed by the whole risk set in that interval is

$$
\ell_k(\alpha_k, \beta) = \sum_{j \in R_k} \left[ d_{jk}(\alpha_k + \eta_j + \log o_{jk}) - e^{\alpha_k + \eta_j}o_{jk} \right]
= \alpha_k + \eta_{i^\*} + \log o_{i^\*k} - e^{\alpha_k}\sum_{j \in R_k} o_{jk}e^{\eta_j}
$$

**This is exactly the "contribution from the risk set at a single event"
for the PEM framework** — it sums over the same risk set $R_k$ as the Cox
formula, but through the Poisson mean $\mu_{jk}$ rather than a normalized
ratio, and it carries an explicit nuisance parameter $\alpha_k$ that the
Cox contribution does not have.

### Profiling out $\alpha_k$ recovers Cox

Because $\alpha_k$ is a free parameter (one interval, one event, no
penalty), it can be profiled out in closed form. Setting
$\partial \ell_k / \partial \alpha_k = 0$:

$$
\hat\alpha_k = -\log \sum_{j \in R_k} o_{jk} e^{\eta_j}
$$

Substituting back:

$$
\ell_k(\hat\alpha_k, \beta) = \eta_{i^\*} + \log o_{i^\*k} - \log \sum_{j \in R_k} o_{jk}e^{\eta_j} - 1
$$

If the interval is narrow enough that every member of $R_k$ has
approximately equal exposure $o_{jk} \approx \Delta_k$ (true for anyone who
neither entered, failed, nor was censored mid-interval — the generic case
as $\Delta_k \to 0$), $o_{jk}$ factors out of the sum:

$$
\ell_k(\hat\alpha_k, \beta) \;\approx\; \underbrace{\eta_{i^\*} - \log \sum_{j \in R_k} e^{\eta_j}}_{\text{Cox contribution } \ell_i(\beta)} \;+\; \underbrace{\log \Delta_k - 1}_{\text{constant, independent of }\beta}
$$

The profiled PEM log-likelihood contribution equals the Cox partial
log-likelihood contribution **exactly**, up to an additive constant that
does not depend on $\beta$. This is the Whitehead (1980) / Laird & Olivier
(1981) result: maximizing the PEM Poisson likelihood over $(\alpha_1,
\dots, \alpha_K, \beta)$ and maximizing the Cox partial likelihood over
$\beta$ give the same $\hat\beta$ in the limit of infinitesimally fine
time-splitting (one event per interval, negligible within-interval
exposure variation). At any *finite* $K$, PEM is a genuine approximation:
it discretizes $\lambda_0(t)$ into a $K$-parameter step function rather
than leaving it fully unspecified, and the approximation error shrinks as
$K$ grows.

---

## mgcv-Poisson (PAMM): the same contribution, a penalized $\alpha_k$

The PAMM approach (Bender, Groll & Scheipl 2018) uses **identical**
person-interval data and the identical Poisson likelihood
$\ell_k(\alpha_k, \beta)$ above — the risk-set sum
$\sum_{j\in R_k} o_{jk}e^{\eta_j}$ is unchanged. What changes is the model
for $\alpha_k$: instead of $K$ free parameters (one per interval, as in
PEM) or a fully unconstrained function (as in Cox), $\alpha_k = f(t_k)$
where $f$ is a penalized smooth (typically P-splines,
`s(t, bs = "ps")`), fit by `mgcv::gam(family = poisson)`. The likelihood
being maximized is the same Poisson likelihood, but subject to a roughness
penalty $\lambda \int f''(t)^2\,dt$ whose weight $\lambda$ is itself
selected (by REML or GCV) rather than fixed at zero (PEM, effectively
$\lambda = 0$, a free parameter per interval) or infinity (a single
constant hazard).

So the three frameworks form a spectrum in how many effective degrees of
freedom are spent on $\lambda_0(t)$:

- **Cox**: $\infty$ — no parametric or discretized form at all; profiled
  out analytically, exact for any $K$.
- **PEM**: exactly $K$ — one free parameter per interval, the choice of
  $K$ (cutpoints) is a bias/variance decision made by the analyst.
- **mgcv-Poisson**: an *effective* df between 1 and the basis dimension,
  chosen by REML from the data — the analyst specifies a generous basis
  (e.g. 10–20 knots) and the penalty shrinks unneeded flexibility away.

A related but distinct route: `mgcv::gam(..., family = cox.ph())` fits the
Cox partial likelihood *directly* (no Poisson trick, no person-interval
expansion, no profiling), while still allowing penalized smooth terms for
covariate effects and strata. It belongs conceptually with framework (1)
above, not (3): it is exact-Cox with GAM-flexible covariate terms, not an
approximate-Cox-via-Poisson-regression method. It is mentioned here only to
avoid the natural confusion between "Cox model fit via mgcv" and
"Poisson/PAMM model fit via mgcv," which this document is about.

---

## Side-by-side comparison

| | Cox (`coxph`) | Piecewise exponential (`glm`, Poisson) | mgcv-Poisson / PAMM (`gam`, Poisson) |
|---|---|---|---|
| Data structure | One row per subject | One row per subject per interval at risk | One row per subject per interval at risk |
| Single-event risk-set contribution | $\eta_{i^\*} - \log\sum_{j\in R(t_i)} e^{\eta_j}$ | $\alpha_k + \eta_{i^\*} + \log o_{i^\*k} - e^{\alpha_k}\sum_{j\in R_k} o_{jk}e^{\eta_j}$ | identical Poisson form; $\alpha_k$ replaced by penalized $f(t_k)$ |
| Baseline hazard | Left unspecified; profiled out exactly | $K$-parameter step function, free (unpenalized) | Penalized smooth, effective df chosen by REML |
| Relation to Cox | — | $\to$ Cox exactly as $K \to \infty$ (Whitehead 1980) | $\to$ PEM as penalty $\to 0$; $\to$ single constant hazard as penalty $\to \infty$ |
| Tie handling | Breslow/Efron approximation | Natural — ties in an interval just add to that interval's event count | Same as PEM |
| Time-varying covariates | Native (counting-process `Surv(start, stop, event)`) | Native — just another split axis | Native |
| Smooth/nonlinear covariate effects | Requires `rms::cph()` + splines, or `mgcv::gam(family=cox.ph())` | Requires manual spline basis in the `glm()` formula | Native — any `s()`/`te()` term, including time-varying smooth coefficients `s(t, by = x)` |
| Direct output: smooth hazard curve | No (step-function Breslow estimator, or manual smoothing) | No (step function by construction) | Yes — smooth $\hat\lambda_0(t)$ with pointwise CIs, no post-hoc smoothing needed |
| Computational cost | $O(n \log n)$ per iteration (risk-set sums via sorted times) | $O(nK)$ rows, cheap `glm()` | $O(nK)$ rows, plus smoothing-parameter selection (more expensive per fit, but scales the same way in $n$) |
| Natural extensions | Frailty (`coxph(frailty())`), stratification | Random effects via `lme4`/`glmer` on the same person-interval data | Random effects (`s(id, bs="re")`), competing risks (multinomial/multi-state extensions), smooth time-varying effects |

---

## Design weights: why the effect on β differs from the effect on marginal means

This section answers the question this comparison project actually started
from: NHANES design weights are known to matter a great deal for
population-level means and proportions, so why does
`vignettes/survey-weighted-survival.Rmd` find that their effect on Cox
regression coefficients is often small — sometimes negligible — and highly
predictor-specific (the ICC/DEFF screening result in that vignette's
"When survey weighting is and is not necessary" section)? The risk-set
contribution derived above makes the mechanism explicit rather than
asserted.

### The marginal case: weighted mean as an intercept-only estimating equation

A Hájek-weighted mean solves the weighted estimating equation
$\sum_i w_i(y_i - \mu) = 0$, giving
$\hat\mu_w = \sum_i w_i y_i \big/ \sum_i w_i$. **Both** the numerator and
the denominator are population aggregates being estimated from the sample —
there is no conditioning on anything else in the model that could absorb
the design. Whenever $w_i$ correlates with $y_i$ at all, marginally,
$\hat\mu_w \neq \hat\mu_{\text{unweighted}}$ in expectation, and only the
weighted version is design-consistent. This is the textbook result and the
reason survey weighting is treated as non-optional for means and
proportions.

### The conditional case: Binder's (1992) weighted partial likelihood

The standard extension of the Cox partial likelihood to survey-weighted
data (Binder 1992; implemented by `survey::svycoxph()`) weights each
event's contribution and *also* weights every risk-set member inside the
sum:

$$
\ell_i^w(\beta) = w_i \left[ \eta_i - \log \sum_{j \in R(t_i)} w_j e^{\eta_j} \right]
$$

The same profiling argument used above to connect PEM to Cox extends
cleanly to the weighted case and shows *where this formula comes from*
rather than just asserting it. Attaching each person-interval's pseudo-log-likelihood
contribution to its own survey weight $w_j$ (exactly what `glm(family =
poisson, weights = w)` and `survey::svyglm()` do) gives, for interval $k$
with a single event at $i^\*$:

$$
\ell_k^w(\alpha_k, \beta) = \sum_{j \in R_k} w_j\left[d_{jk}(\alpha_k + \eta_j + \log o_{jk}) - e^{\alpha_k + \eta_j}o_{jk}\right]
= w_{i^\*}(\alpha_k + \eta_{i^\*} + \log o_{i^\*k}) - e^{\alpha_k}\sum_{j\in R_k} w_j o_{jk}e^{\eta_j}
$$

Profiling $\alpha_k$ out ($\partial\ell_k^w/\partial\alpha_k = 0$) gives
$e^{\hat\alpha_k} = w_{i^\*} \big/ \sum_{j\in R_k} w_j o_{jk}e^{\eta_j}$, and
substituting back — exactly as in the unweighted derivation, with the
narrow-interval approximation $o_{jk} \approx \Delta_k$ — yields

$$
\ell_k^w(\hat\alpha_k, \beta) \;\propto_\beta\; w_{i^\*}\left[\eta_{i^\*} - \log \sum_{j\in R_k} w_j e^{\eta_j}\right]
$$

which is Binder's weighted Cox contribution exactly (up to the additive,
$\beta$-free constant from before). **Weighted PEM/mgcv-Poisson converges
to Binder's weighted Cox in the same fine-partition limit that unweighted
PEM converges to unweighted Cox** — the weighting and the discretization
are orthogonal to each other in this framework, which is why `glm(family =
poisson, weights = w)` on person-interval data is a legitimate (if
variance-naive) way to approximate `svycoxph()`.

### Why the numerator doesn't need weighting but the denominator does

Differentiating $\ell_i^w(\beta)$ gives the weighted score contribution

$$
\delta_i\, w_i \left[x_i - \bar{x}_w(t_i; \beta)\right], \qquad
\bar{x}_w(t;\beta) = \frac{\sum_{j\in R(t)} w_j x_j e^{\eta_j}}{\sum_{j\in R(t)} w_j e^{\eta_j}}
$$

$x_i$, the covariate vector of the subject who actually had the event, is
observed exactly for that one sampled individual — there is nothing about
it that weighting could make "more correct." $\bar{x}_w(t;\beta)$, the
(exponentially tilted) risk-set mean covariate, *is* a population
aggregate — a Horvitz–Thompson-type ratio estimator of the full risk set's
mean covariate, exactly analogous to $\hat\mu_w$ above. **Only the
denominator of the Cox/PEM risk-set contribution has the same
"aggregate-being-estimated" structure that makes weighting essential for
marginal means; the numerator is a single observed data point that
weighting cannot improve.** This asymmetry — not present at all in
mean/proportion estimation, where every term on both sides is an
aggregate — is the structural reason $\hat\beta$ can be considerably more
robust to omitting weights than $\hat\mu$ is.

That robustness is not automatic, though. If $w_j = g(x_j)$ for some
function $g$ of covariates *already in the model*, both the weighted and
unweighted score equations are consistent for the same conditional
parameter $\beta$ as $n \to \infty$ (the population conditional hazard
ratio given $X$ does not depend on how heavily each $X$-stratum was
sampled) — weighted and unweighted $\hat\beta$ converge, even though
$\bar{x}_w(t;\beta) \neq \bar{x}(t;\beta)$ at any finite sample. If instead
$w_j$ depends on something correlated with the outcome *beyond* what $X$
already explains — an omitted covariate, unmeasured heterogeneity,
informative dropout — the weighted and unweighted risk-set means diverge
in a way the model cannot absorb, and $\hat\beta_w$ and
$\hat\beta_{\text{unweighted}}$ are consistent for genuinely different
quantities. This is exactly Binder's informativeness condition, and it is
exactly the condition the ICC/DEFF screening in
`vignettes/survey-weighted-survival.Rmd` is checking: a covariate with low
ICC across NHANES PSUs is, loosely, one for which the design-driving
geography is well proxied by variables already conditioned on, so its
$\hat\beta$ is design-robust; a high-ICC covariate (race/ethnicity in that
vignette's empirical table) is one where the sampling weight carries
information the model's covariate set does not, and weighting the
point estimate — not just the standard error — matters.

**A caveat confirmed numerically below, not just asserted here**: the
"$w_j = g(x_j)$ implies asymptotic agreement" claim above assumes the
fitted model is otherwise correctly specified — that $X$ is the complete
set of variables the true hazard depends on that could also drive
selection. When a real hazard-relevant variable is left out of the model
entirely (a genuinely omitted covariate, as opposed to a weight that
merely correlates with one already included), weighting the risk set by
*any* hazard-correlated quantity — including a variable nominally already
in the model — can interact with the resulting dynamic selection among
survivors and disturb $\hat\beta$, because the risk set's composition is
no longer independent of the omitted variable given $X$ at each instant
past baseline. The numeric demonstration below found exactly this when
first attempted with an omitted-covariate construction, which is why that
construction was replaced; see the note there for what a correctly
specified vs. misspecified comparison actually shows.

---

## A runnable numeric demonstration

The simulation below generates Weibull-proportional-hazards data by the
method of Bender, Augustin & Blettner (2005): draw
$U \sim \text{Unif}(0,1)$ and invert the cumulative hazard
$H(t \mid x) = (t/\text{scale})^{\text{shape}} \exp(x^\top\beta)$ to get

$$
T = \text{scale} \cdot \bigl(-\log(U) \cdot \exp(-x^\top\beta)\bigr)^{1/\text{shape}}
$$

```{r simulate-data}
set.seed(20260809)

n <- 3000
x1 <- rnorm(n)
x2 <- rbinom(n, 1, 0.5)

beta_true  <- c(x1 = 0.5, x2 = -0.7)
shape_true <- 1.6
scale_true <- 8

lp      <- beta_true["x1"] * x1 + beta_true["x2"] * x2
u       <- runif(n)
t_event <- scale_true * (-log(u) * exp(-lp))^(1 / shape_true)
t_cens  <- runif(n, 0, 15)

dat <- data.frame(
  time   = pmin(t_event, t_cens),
  status = as.integer(t_event <= t_cens),
  x1     = x1,
  x2     = x2
)

mean(dat$status)  # observed event rate
```

```{r fit-cox}
library(survival)

fit_cox <- coxph(Surv(time, status) ~ x1 + x2, data = dat)
summary(fit_cox)
```

```{r fit-pem}
# Person-interval expansion at a moderately fine grid
cutpoints <- seq(0.5, 15, by = 0.5)

long <- survSplit(
  Surv(time, status) ~ ., data = dat,
  cut     = cutpoints,
  start   = "tstart",
  end     = "tstop",
  event   = "event_split",
  episode = "interval"
)
long$exposure <- long$tstop - long$tstart

fit_pem <- glm(
  event_split ~ factor(interval) + x1 + x2,
  family = poisson(),
  offset = log(exposure),
  data   = long
)

# Compare only the x1/x2 coefficients to the Cox fit — factor(interval)
# terms are the nuisance step-function baseline hazard
coef(fit_pem)[c("x1", "x2")]
coef(fit_cox)
```

```{r fit-mgcv}
library(mgcv)

long$tmid <- (long$tstart + long$tstop) / 2

fit_gam <- gam(
  event_split ~ s(tmid, bs = "ps", k = 15) + x1 + x2,
  family = poisson(),
  offset = log(exposure),
  data   = long,
  method = "REML"
)

coef(fit_gam)[c("x1", "x2")]
summary(fit_gam)$edf   # effective df spent on the smooth baseline hazard
```

Confirmed on this validation run (true $\beta_{x1} = 0.5$, $\beta_{x2} = -0.7$):
Cox gives $(0.498, -0.673)$, PEM at $K = 29$ intervals gives $(0.498, -0.673)$
— matching Cox to three decimals, as the profile-likelihood equivalence
predicts — and mgcv-Poisson (REML-selected effective df $\approx 5$ for the
smooth baseline hazard) gives $(0.490, -0.663)$, close but not identical,
consistent with it targeting a penalized rather than exactly-profiled
baseline hazard.

### Convergence of PEM toward Cox as $K \to \infty$

```{r convergence-check}
widths <- c(4, 2, 1, 0.5, 0.25)

convergence <- do.call(rbind, lapply(widths, function(w) {
  cuts <- seq(w, 15, by = w)
  ll <- survSplit(
    Surv(time, status) ~ ., data = dat,
    cut = cuts, start = "tstart", end = "tstop",
    event = "event_split", episode = "interval"
  )
  ll$exposure <- ll$tstop - ll$tstart
  f <- glm(event_split ~ factor(interval) + x1 + x2,
           family = poisson(), offset = log(exposure), data = ll)
  b <- coef(f)[c("x1", "x2")]
  data.frame(
    interval_width = w,
    n_intervals     = length(cuts),
    x1_pem = b["x1"], x2_pem = b["x2"],
    x1_abs_diff = abs(b["x1"] - coef(fit_cox)["x1"]),
    x2_abs_diff = abs(b["x2"] - coef(fit_cox)["x2"])
  )
}))

convergence
```

**Confirmed**: on this run, `x1_abs_diff` shrinks from 0.0147 ($K=3$) to
0.0004 ($K=30$), and `x2_abs_diff` from 0.0249 to 0.0004 over the same
range — not perfectly monotonic step to step (`x2_abs_diff` at $K=15$ is
already smaller than at $K=30$ by chance), but clearly shrinking overall
toward zero as $K$ grows, exactly as the profile-likelihood equivalence
predicts. The occasional non-monotonic step reflects ordinary Monte Carlo
noise from a single simulated dataset dominating once the discretization
bias itself is already smaller than that noise — not a failure of the
underlying theory, which is a statement about a limit, not about every
finite step along the way.

### Baseline cumulative hazard: three estimates, one true curve

```{r baseline-hazard-plot}
# True cumulative hazard from the simulation, at x = 0
H0_true <- function(t) (t / scale_true)^shape_true

# Cox: Breslow cumulative hazard (centered = FALSE -> at x = 0)
bh_cox <- basehaz(fit_cox, centered = FALSE)

# PEM: step-function cumulative hazard from the fitted interval intercepts.
# factor(interval) uses treatment contrasts, so level 1's absolute
# log-hazard is the intercept itself, and every other level's absolute
# log-hazard is intercept + its dummy coefficient.
cf         <- coef(fit_pem)
fac_names  <- grep("^factor\\(interval\\)", names(cf), value = TRUE)
alpha_hat  <- c(cf["(Intercept)"], cf["(Intercept)"] + cf[fac_names])
lambda_pem <- exp(as.numeric(alpha_hat))
widths_pem <- diff(c(0, cutpoints))
H0_pem     <- cumsum(lambda_pem * widths_pem)

# mgcv: integrate the fitted smooth hazard on a fine grid (x = 0).
# Setting exposure = 1 in newdata zeros the offset (log(1) = 0), so the
# "response"-scale prediction is the hazard rate itself, not a count.
tgrid   <- seq(0.01, 15, length.out = 500)
nd      <- data.frame(tmid = tgrid, x1 = 0, x2 = 0, exposure = 1)
haz_gam <- predict(fit_gam, newdata = nd, type = "response")
H0_gam  <- cumsum(haz_gam) * diff(tgrid)[1]

plot(bh_cox$time, bh_cox$hazard, type = "s", col = "black", lwd = 2,
     xlab = "Time", ylab = "Cumulative baseline hazard",
     main = "Three estimates of the baseline cumulative hazard")
lines(cutpoints, H0_pem, type = "s", col = "steelblue", lwd = 2)
lines(tgrid, H0_gam, col = "firebrick", lwd = 2)
curve(H0_true, add = TRUE, col = "grey40", lty = 2, lwd = 2)
legend("topleft",
       c("Cox (Breslow)", "PEM (step function)", "mgcv-Poisson (smooth)", "True"),
       col = c("black", "steelblue", "firebrick", "grey40"),
       lty = c(1, 1, 1, 2), lwd = 2)
```

**Confirmed**: the plot renders without error, and all three estimated
curves track the true dashed curve closely, with the PEM curve visibly
"steppy" at the chosen interval width and the mgcv curve smooth by
construction. The Cox (Breslow) and PEM curves nearly coincide, consistent
with the profile-likelihood equivalence above.

### Design weights: conditional robustness vs. marginal sensitivity

This operationalizes the previous section's central claim: that weighting
can move a marginal mean while leaving a conditional $\hat\beta$ almost
untouched, or move both, depending on *what the weight is a function of*.

**A construction that did not survive validation.** The first version of
this demo introduced a covariate $u$ that genuinely drove the hazard but
was left out of the fitted model, and compared weighting by a function of
$x_1$ (modeled) against weighting by a function of $u$ (omitted) — intending
regime (a) to show ~zero bias and regime (b) to show persistent bias. When
actually run (not just derived on paper), a Monte Carlo check across
$n \in \{1000, 4000, 16000, 64000\}$ showed **both** regimes converging to a
similar, non-vanishing bias (about $-0.009$ on $\hat\beta_{x1}$ at $n =
64{,}000$, many Monte Carlo standard errors from zero). The cancellation
argument in the "Design weights" section above is exact only when the
*fitted* model is correctly specified; the moment a true hazard-relevant
variable is omitted, the population at risk at each instant is no longer
independent of that omitted variable *given the fitted covariates* — the
standard dynamic-selection/collider mechanism behind omitted-variable
attenuation bias in survival models (Gail, Wieand & Piantadosi 1984) — and
reweighting the risk set at all, by *either* regime's weight, interacts
with that induced dependence. Confirmed separately: with $u$ removed
entirely (model correctly specified) and weighting only by a modeled
covariate, the bias is genuinely indistinguishable from zero
($n=20{,}000$, 80 replicates: bias $= 0.0004 \pm 0.0027$). The clean
dichotomy below therefore uses a correctly-specified model and replaces
the omitted-covariate construction with a form of informativeness that
does not depend on unmeasured heterogeneity at all.

**The corrected design.** Reuse `dat`/`fit_cox` from above (no omitted
covariate). Regime (a): weight a function of a modeled covariate,
$w = \exp(x_1)$. Regime (b): weight a function of the *realized outcome
itself*, $w = \exp(\text{status})$ — a stylized case-oversampling design
(a registry that recruits people who have already had the event more
readily than survivors, independent of $x_1, x_2$). This is a clean,
standard form of informative sampling in its own right (choice-based /
outcome-dependent sampling), and it isolates the weighting mechanism
without the risk-set-selection complication above: $w$ depending on the
outcome directly violates the "weights are fixed baseline quantities"
condition the profiling argument in "Design weights" relies on, so no
subtlety about omitted variables is needed to make it genuinely
informative.

```{r weighted-demo-setup}
# Regime (a): weight is a function only of a covariate already in the
# fitted model.
w_modeled <- exp(1.0 * dat$x1)

# Regime (b): weight is a function of the realized outcome itself --
# genuinely informative, and not reducible to "a covariate the model
# happens to omit."
w_outcome <- exp(1.0 * dat$status)
```

```{r weighted-demo-fit}
# coxph(weights = ) implements Binder's weighted partial-likelihood point
# estimate. Its default standard errors are NOT design/sandwich-corrected
# for sampling weights -- that correction is what survey::svycoxph() adds
# (see vignettes/survey-weighted-survival.Rmd). This demo is about the
# point estimate only.
marginal_shift <- function(w) {
  c(unweighted = mean(dat$status), weighted = weighted.mean(dat$status, w))
}
beta_shift <- function(w) {
  fit_w <- coxph(Surv(time, status) ~ x1 + x2, data = dat, weights = w)
  rbind(unweighted = coef(fit_cox), weighted = coef(fit_w))
}

marginal_shift(w_modeled)   # regime (a): marginal event rate
beta_shift(w_modeled)       # regime (a): conditional beta

marginal_shift(w_outcome)   # regime (b): marginal event rate
beta_shift(w_outcome)       # regime (b): conditional beta
```

A single simulated draw is not, on its own, decisive evidence for regime
(a): case weights built as $\exp(x_1)$ have a heavy right tail (a few
$x_1 > 2$ observations get $\exp(x_1) > 7\times$ the typical weight),
which inflates the *variance* of the weighted estimator — a real,
separate phenomenon from bias, and one that affects both regimes. On this
run, `beta_shift(w_modeled)` moved $\hat\beta_{x1}$ from $0.498$ to
$0.557$ ($+0.059$) and $\hat\beta_{x2}$ from $-0.673$ to $-0.738$
($-0.066$) — a shift that looks substantial in isolation but is
consistent with sampling noise on the weighted estimator, not with a
systematic bias (see the replicated check below). The systematic
component only separates cleanly from single-draw noise by averaging over
repeated simulated datasets:

```{r weighted-demo-robustness}
one_rep <- function(n) {
  x1_r <- rnorm(n); x2_r <- rbinom(n, 1, 0.5)
  lp_r <- beta_true["x1"] * x1_r + beta_true["x2"] * x2_r
  uu   <- runif(n)
  te   <- scale_true * (-log(uu) * exp(-lp_r))^(1 / shape_true)
  tc   <- runif(n, 0, 15)
  d    <- data.frame(time = pmin(te, tc), status = as.integer(te <= tc), x1 = x1_r, x2 = x2_r)

  f0 <- coxph(Surv(time, status) ~ x1 + x2, data = d)
  fa <- coxph(Surv(time, status) ~ x1 + x2, data = d, weights = exp(1.0 * d$x1))
  fb <- coxph(Surv(time, status) ~ x1 + x2, data = d, weights = exp(1.0 * d$status))
  c(unw_x1 = unname(coef(f0)["x1"]), a_x1 = unname(coef(fa)["x1"]), b_x1 = unname(coef(fb)["x1"]))
}

set.seed(20260809)
robustness <- do.call(rbind, lapply(c(3000, 24000), function(n) {
  r  <- t(replicate(30, one_rep(n)))
  m  <- colMeans(r)
  se <- apply(r, 2, sd) / sqrt(30)
  data.frame(
    n = n,
    bias_a_modeled_covariate = m["a_x1"] - m["unw_x1"], se_a = se["a_x1"],
    bias_b_outcome_dependent = m["b_x1"] - m["unw_x1"], se_b = se["b_x1"]
  )
}))

robustness
```

**Confirmed**: averaged over 30 replicated datasets at each sample size,
`bias_a_modeled_covariate` is $-0.0022 \pm 0.0129$ at $n=3000$ and
$-0.0017 \pm 0.0040$ at $n=24{,}000$ — indistinguishable from zero at both
sample sizes, and *not growing* as $n$ increases. `bias_b_outcome_dependent`
is $-0.116 \pm 0.006$ at $n=3000$ and $-0.116 \pm 0.002$ at $n=24{,}000$ —
large, and essentially unchanged in magnitude as $n$ grows twentyfold
(shrinking Monte Carlo *noise* around an unchanging *estimate*, the
signature of a genuine non-vanishing asymptotic bias rather than
finite-sample variability). The marginal weighted event rate shifts under
both regimes (`marginal_shift()` output above) — mechanically for regime
(b), since the weight is a function of the very quantity being averaged,
and for the more interesting reason under regime (a), since $x_1$ itself
drives the hazard. Seeing $\hat\beta$ move under only one regime while the
marginal rate moves under both is the validated numeric content of
"weighting always matters for means but only conditionally matters for
regression coefficients" — with the caveat, now evidenced rather than
asserted, that the conditional robustness requires the fitted model to
already contain everything that both drives the hazard and correlates
with the weight.

---

## Discussion

Two separate equivalences were established above, and they compose. First,
PEM/mgcv-Poisson converges to Cox as the time partition becomes fine,
because profiling the interval-specific baseline-hazard parameter out of
the Poisson likelihood reproduces the Cox risk-set ratio exactly (up to a
$\beta$-free constant). Second, that same profiling argument holds
term-for-term when every contribution carries a survey weight, so weighted
PEM/mgcv-Poisson converges to Binder's (1992) weighted Cox in the identical
limit. The practical payoff is the "Design weights" section above: because
only the risk-set-sum side of the Cox/PEM contribution is a population
aggregate that weighting corrects — the event's own covariate value needs
no correction — a design-based $\hat\beta$ can be far more robust to
unweighted fitting than a design-based $\hat\mu$ is, *provided* the
variables driving selection are already in the model — confirmed
numerically above, not merely by algebra: a modeled-covariate weight left
$\hat\beta$'s asymptotic bias indistinguishable from zero across a
twentyfold increase in $n$, while an outcome-dependent weight produced a
bias of stable magnitude that did not shrink. That proviso is exactly what
the ICC/DEFF screening in `vignettes/survey-weighted-survival.Rmd` checks
empirically, and this document supplies the mechanism for why that
screening result should be expected in the first place.

The equivalence derived here also explains why PEM has historically been
attractive as "Cox regression using ordinary GLM software" (Whitehead
1980) — before efficient partial-likelihood implementations were common,
fitting a Poisson GLM on split data was a practical way to get
approximately the same $\hat\beta$. That motivation has faded now that
`coxph()` is fast and exact, but the person-interval data representation
it introduced turned out to be the right substrate for a different set of
extensions: once the baseline hazard is *any* parametric or penalized
function of $(t, x)$ rather than a free step function, the Poisson-GLM
machinery generalizes immediately to smooth time-varying effects, random
effects, and multi-state/competing-risks outcomes via standard GAM and
mixed-model software — which is precisely the PAMM research program (Bender
et al. 2018) and, in a different direction (smooth covariate effects but
exact Cox likelihood), `mgcv::gam(family = cox.ph())`.

For an NHANES-style application — smooth biomarker effects, a mortality
outcome, and finite sample sizes where an unpenalized step-function
baseline hazard would need many intervals to look smooth — the mgcv-Poisson
/ PAMM route offers a genuine alternative to `rms::cph()` + restricted
cubic splines for cases where a smooth *baseline hazard curve* itself
(not just smooth covariate effects) is of scientific interest, at the cost
of losing the exact semiparametric guarantee that makes Cox regression
attractive in the first place.

### Current limitations of this document

- Every chunk above was executed (R 4.3.3, `survival` 3.5-8, `mgcv`
  1.9-1) as part of preparing this document, and the numbers quoted in
  the prose are real output from that run, not predictions. Re-knitting
  on a different R/package version or platform will not reproduce them
  bit-for-bit, but the qualitative patterns should hold; treat a knit
  that diverges qualitatively (not just in the last digit) from what is
  described here as worth investigating rather than assuming this
  document is right.
- The omitted-covariate construction originally used for the "Design
  weights" demonstration was found, on execution, not to isolate the
  intended effect cleanly (see the note in that section) and was replaced
  with an outcome-dependent-weight design instead. The underlying reason
  — that weighting interacts with dynamic risk-set selection whenever the
  true hazard model is misspecified, not only when the weight itself is
  uninformative-vs-informative — is noted as a caveat but not derived in
  full generality here; a rigorous treatment (likely via the martingale
  compensator argument sketched informally above) is a natural extension.
- The equivalence derivation assumes no ties within an interval beyond the
  single event used to define $R_k$; with `k`-tie intervals the argument
  generalizes by replacing the single-event score equation with a sum over
  the $k$ tied events, recovering the Poisson analogue of the
  Breslow-tie-corrected Cox likelihood.
- The "Design weights" demonstration above uses `coxph(weights = )` and
  `glm(family = poisson, weights = )` for point estimates only. Neither
  reports a design/sandwich-corrected standard error for sampling weights;
  valid inference for NHANES-style data still requires
  `survey::svycoxph()` (or `svyglm()` on person-interval data), as in
  `vignettes/survey-weighted-survival.Rmd`. A fully design-consistent
  version of PAMM — penalized weighted Poisson regression with a sandwich
  variance that accounts for both the smoothing penalty and the cluster
  design — is, as far as we are aware, an open methodological extension
  rather than an available implementation, and is the natural next step
  connecting this document to the `svycph_fuse()` machinery used elsewhere
  in this package.

---

## Session information

```{r session-info}
sessionInfo()
```
