---
title: "Comparing Survival Frameworks: Cox, Piecewise Exponential, and mgcv-Poisson"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Survival Frameworks: Cox, Piecewise Exponential, and mgcv-Poisson}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

> **Note on chunk evaluation.** This vignette is fully self-contained — it
> simulates its own data and requires no network access or external files —
> but chunks are set `eval = FALSE` because it was authored in an environment
> without an R installation available to verify execution. Set
> `knitr::opts_chunk$set(eval = TRUE)` (or knit normally, which honors chunk
> options as written unless overridden) once you have confirmed the code runs
> cleanly, and fold the results back into the prose before treating any
> numbers below as real. Everything under "Expected pattern" is a
> qualitative prediction from the theory in this document, not an executed
> result.

## Motivation

Three common frameworks for estimating a covariate-adjusted hazard produce
numerically related, and in one limiting case *identical*, answers:

1. **Cox proportional hazards** (`survival::coxph()`) — semiparametric,
   partial likelihood, baseline hazard left completely unspecified.
2. **Piecewise exponential model (PEM)** — the time axis is cut into
   intervals, the hazard is assumed constant within each interval, and the
   model is fit as a Poisson GLM on person-interval records
   (`stats::glm(family = poisson)`).
3. **mgcv-Poisson / PAMM** (piecewise-exponential *additive* model,
   Bender, Groll & Scheipl 2018) — the same person-interval data as (2), but
   the interval-specific log-hazards are replaced by a penalized smooth
   function of time, fit with `mgcv::gam(family = poisson)`.

These are usually presented as three unrelated modeling choices. They are
not: (2) is a discretized, computationally cheap approximation to (1), and
(3) is what you get when you stop forcing the discretization to be a step
function and let a penalized spline choose its own effective smoothness.
The clearest way to see the relationship is to look at what each framework
contributes to its likelihood **from the risk set present at a single event
time** — the atomic unit every one of these models is built from.

This document derives that contribution for all three frameworks side by
side, shows the algebraic equivalence between (1) and (2) in the limit of
fine time-splitting, and gives a runnable simulation comparing all three
numerically.

### Key references

- Whitehead, J. (1980). Fitting Cox's regression model to survival data
  using GLIM. *Applied Statistics*, 29(3), 268–275.
- Laird, N. & Olivier, D. (1981). Covariance analysis of censored survival
  data using log-linear analysis techniques. *JASA*, 76(374), 231–240.
- Holford, T.R. (1980). The analysis of rates and of survivorship using
  log-linear models. *Biometrics*, 36(2), 299–305.
- Bender, A., Groll, A. & Scheipl, F. (2018). A generalized additive model
  approach to time-to-event analysis. *Statistical Modelling*, 18(3-4),
  299–321. [PAMMs; see also the `pammtools` package]
- Wood, S.N. (2017). *Generalized Additive Models: An Introduction with R*,
  2nd ed. CRC Press. [`mgcv`]
- Therneau, T.M. & Grambsch, P.M. (2000). *Modeling Survival Data: Extending
  the Cox Model*. Springer.
- Bender, R., Augustin, T. & Blettner, M. (2005). Generating survival times
  to simulate Cox proportional hazards models. *Statistics in Medicine*,
  24(11), 1713–1723. [the simulation scheme used below]

---

## Notation

For subject $j$ with covariate vector $x_j$ and linear predictor
$\eta_j = x_j^\top \beta$, let $R(t)$ denote the risk set at time $t$ — the
set of subjects still under observation (not yet failed or censored) just
before $t$. An event occurs at time $t_i$ for subject $i^\*(t_i) \in R(t_i)$.

---

## Cox: the partial-likelihood contribution

Ignoring ties, the Cox partial likelihood contribution at event time
$t_i$ is the now-familiar risk-set ratio:

$$
L_i(\beta) = \frac{\exp(\eta_{i^\*})}{\sum_{j \in R(t_i)} \exp(\eta_j)},
\qquad
\ell_i(\beta) = \eta_{i^\*} - \log \sum_{j \in R(t_i)} \exp(\eta_j)
$$

The baseline hazard $\lambda_0(t)$ never appears: it cancels out of the
ratio entirely. This is the sense in which Cox regression is
semiparametric — $\lambda_0(t)$ is an infinite-dimensional nuisance
parameter that is *profiled out analytically*, not estimated. Ties are
handled by the Breslow or Efron approximation to the exact discrete-risk-set
likelihood; `coxph()` uses Efron by default.

---

## Piecewise exponential: the Poisson-trick contribution

Partition the time axis into intervals with cutpoints
$0 = c_0 < c_1 < \dots < c_K$. Every subject contributes one
person-interval record for each interval they were at risk in. For
subject $j$ in interval $k$: exposure time $o_{jk}$ (time spent at risk in
that interval), event indicator $d_{jk} \in \{0, 1\}$ (1 only if the event
happened in that interval), and assumed constant hazard within the
interval, $\lambda_k(x_j) = \lambda_{0k}\exp(\eta_j)$. Then
$d_{jk} \sim \text{Poisson}(\mu_{jk})$ with
$\mu_{jk} = o_{jk}\lambda_{0k}\exp(\eta_j) = \exp(\log o_{jk} + \alpha_k + \eta_j)$,
i.e. a Poisson GLM with an offset of $\log o_{jk}$ and a free intercept
$\alpha_k = \log \lambda_{0k}$ per interval.

For an interval $k$ containing exactly one event, at subject $i^\*$, the
Poisson log-likelihood contributed by the whole risk set in that interval is

$$
\ell_k(\alpha_k, \beta) = \sum_{j \in R_k} \left[ d_{jk}(\alpha_k + \eta_j + \log o_{jk}) - e^{\alpha_k + \eta_j}o_{jk} \right]
= \alpha_k + \eta_{i^\*} + \log o_{i^\*k} - e^{\alpha_k}\sum_{j \in R_k} o_{jk}e^{\eta_j}
$$

**This is exactly the "contribution from the risk set at a single event"
for the PEM framework** — it sums over the same risk set $R_k$ as the Cox
formula, but through the Poisson mean $\mu_{jk}$ rather than a normalized
ratio, and it carries an explicit nuisance parameter $\alpha_k$ that the
Cox contribution does not have.

### Profiling out $\alpha_k$ recovers Cox

Because $\alpha_k$ is a free parameter (one interval, one event, no
penalty), it can be profiled out in closed form. Setting
$\partial \ell_k / \partial \alpha_k = 0$:

$$
\hat\alpha_k = -\log \sum_{j \in R_k} o_{jk} e^{\eta_j}
$$

Substituting back:

$$
\ell_k(\hat\alpha_k, \beta) = \eta_{i^\*} + \log o_{i^\*k} - \log \sum_{j \in R_k} o_{jk}e^{\eta_j} - 1
$$

If the interval is narrow enough that every member of $R_k$ has
approximately equal exposure $o_{jk} \approx \Delta_k$ (true for anyone who
neither entered, failed, nor was censored mid-interval — the generic case
as $\Delta_k \to 0$), $o_{jk}$ factors out of the sum:

$$
\ell_k(\hat\alpha_k, \beta) \;\approx\; \underbrace{\eta_{i^\*} - \log \sum_{j \in R_k} e^{\eta_j}}_{\text{Cox contribution } \ell_i(\beta)} \;+\; \underbrace{\log \Delta_k - 1}_{\text{constant, independent of }\beta}
$$

The profiled PEM log-likelihood contribution equals the Cox partial
log-likelihood contribution **exactly**, up to an additive constant that
does not depend on $\beta$. This is the Whitehead (1980) / Laird & Olivier
(1981) result: maximizing the PEM Poisson likelihood over $(\alpha_1,
\dots, \alpha_K, \beta)$ and maximizing the Cox partial likelihood over
$\beta$ give the same $\hat\beta$ in the limit of infinitesimally fine
time-splitting (one event per interval, negligible within-interval
exposure variation). At any *finite* $K$, PEM is a genuine approximation:
it discretizes $\lambda_0(t)$ into a $K$-parameter step function rather
than leaving it fully unspecified, and the approximation error shrinks as
$K$ grows.

---

## mgcv-Poisson (PAMM): the same contribution, a penalized $\alpha_k$

The PAMM approach (Bender, Groll & Scheipl 2018) uses **identical**
person-interval data and the identical Poisson likelihood
$\ell_k(\alpha_k, \beta)$ above — the risk-set sum
$\sum_{j\in R_k} o_{jk}e^{\eta_j}$ is unchanged. What changes is the model
for $\alpha_k$: instead of $K$ free parameters (one per interval, as in
PEM) or a fully unconstrained function (as in Cox), $\alpha_k = f(t_k)$
where $f$ is a penalized smooth (typically P-splines,
`s(t, bs = "ps")`), fit by `mgcv::gam(family = poisson)`. The likelihood
being maximized is the same Poisson likelihood, but subject to a roughness
penalty $\lambda \int f''(t)^2\,dt$ whose weight $\lambda$ is itself
selected (by REML or GCV) rather than fixed at zero (PEM, effectively
$\lambda = 0$, a free parameter per interval) or infinity (a single
constant hazard).

So the three frameworks form a spectrum in how many effective degrees of
freedom are spent on $\lambda_0(t)$:

- **Cox**: $\infty$ — no parametric or discretized form at all; profiled
  out analytically, exact for any $K$.
- **PEM**: exactly $K$ — one free parameter per interval, the choice of
  $K$ (cutpoints) is a bias/variance decision made by the analyst.
- **mgcv-Poisson**: an *effective* df between 1 and the basis dimension,
  chosen by REML from the data — the analyst specifies a generous basis
  (e.g. 10–20 knots) and the penalty shrinks unneeded flexibility away.

A related but distinct route: `mgcv::gam(..., family = cox.ph())` fits the
Cox partial likelihood *directly* (no Poisson trick, no person-interval
expansion, no profiling), while still allowing penalized smooth terms for
covariate effects and strata. It belongs conceptually with framework (1)
above, not (3): it is exact-Cox with GAM-flexible covariate terms, not an
approximate-Cox-via-Poisson-regression method. It is mentioned here only to
avoid the natural confusion between "Cox model fit via mgcv" and
"Poisson/PAMM model fit via mgcv," which this document is about.

---

## Side-by-side comparison

| | Cox (`coxph`) | Piecewise exponential (`glm`, Poisson) | mgcv-Poisson / PAMM (`gam`, Poisson) |
|---|---|---|---|
| Data structure | One row per subject | One row per subject per interval at risk | One row per subject per interval at risk |
| Single-event risk-set contribution | $\eta_{i^\*} - \log\sum_{j\in R(t_i)} e^{\eta_j}$ | $\alpha_k + \eta_{i^\*} + \log o_{i^\*k} - e^{\alpha_k}\sum_{j\in R_k} o_{jk}e^{\eta_j}$ | identical Poisson form; $\alpha_k$ replaced by penalized $f(t_k)$ |
| Baseline hazard | Left unspecified; profiled out exactly | $K$-parameter step function, free (unpenalized) | Penalized smooth, effective df chosen by REML |
| Relation to Cox | — | $\to$ Cox exactly as $K \to \infty$ (Whitehead 1980) | $\to$ PEM as penalty $\to 0$; $\to$ single constant hazard as penalty $\to \infty$ |
| Tie handling | Breslow/Efron approximation | Natural — ties in an interval just add to that interval's event count | Same as PEM |
| Time-varying covariates | Native (counting-process `Surv(start, stop, event)`) | Native — just another split axis | Native |
| Smooth/nonlinear covariate effects | Requires `rms::cph()` + splines, or `mgcv::gam(family=cox.ph())` | Requires manual spline basis in the `glm()` formula | Native — any `s()`/`te()` term, including time-varying smooth coefficients `s(t, by = x)` |
| Direct output: smooth hazard curve | No (step-function Breslow estimator, or manual smoothing) | No (step function by construction) | Yes — smooth $\hat\lambda_0(t)$ with pointwise CIs, no post-hoc smoothing needed |
| Computational cost | $O(n \log n)$ per iteration (risk-set sums via sorted times) | $O(nK)$ rows, cheap `glm()` | $O(nK)$ rows, plus smoothing-parameter selection (more expensive per fit, but scales the same way in $n$) |
| Natural extensions | Frailty (`coxph(frailty())`), stratification | Random effects via `lme4`/`glmer` on the same person-interval data | Random effects (`s(id, bs="re")`), competing risks (multinomial/multi-state extensions), smooth time-varying effects |

---

## A runnable numeric demonstration

The simulation below generates Weibull-proportional-hazards data by the
method of Bender, Augustin & Blettner (2005): draw
$U \sim \text{Unif}(0,1)$ and invert the cumulative hazard
$H(t \mid x) = (t/\text{scale})^{\text{shape}} \exp(x^\top\beta)$ to get

$$
T = \text{scale} \cdot \bigl(-\log(U) \cdot \exp(-x^\top\beta)\bigr)^{1/\text{shape}}
$$

```{r simulate-data}
set.seed(20260809)

n <- 3000
x1 <- rnorm(n)
x2 <- rbinom(n, 1, 0.5)

beta_true  <- c(x1 = 0.5, x2 = -0.7)
shape_true <- 1.6
scale_true <- 8

lp      <- beta_true["x1"] * x1 + beta_true["x2"] * x2
u       <- runif(n)
t_event <- scale_true * (-log(u) * exp(-lp))^(1 / shape_true)
t_cens  <- runif(n, 0, 15)

dat <- data.frame(
  time   = pmin(t_event, t_cens),
  status = as.integer(t_event <= t_cens),
  x1     = x1,
  x2     = x2
)

mean(dat$status)  # observed event rate
```

```{r fit-cox}
library(survival)

fit_cox <- coxph(Surv(time, status) ~ x1 + x2, data = dat)
summary(fit_cox)
```

```{r fit-pem}
# Person-interval expansion at a moderately fine grid
cutpoints <- seq(0.5, 15, by = 0.5)

long <- survSplit(
  Surv(time, status) ~ ., data = dat,
  cut     = cutpoints,
  start   = "tstart",
  end     = "tstop",
  event   = "event_split",
  episode = "interval"
)
long$exposure <- long$tstop - long$tstart

fit_pem <- glm(
  event_split ~ factor(interval) + x1 + x2,
  family = poisson(),
  offset = log(exposure),
  data   = long
)

# Compare only the x1/x2 coefficients to the Cox fit — factor(interval)
# terms are the nuisance step-function baseline hazard
coef(fit_pem)[c("x1", "x2")]
coef(fit_cox)
```

```{r fit-mgcv}
library(mgcv)

long$tmid <- (long$tstart + long$tstop) / 2

fit_gam <- gam(
  event_split ~ s(tmid, bs = "ps", k = 15) + x1 + x2,
  family = poisson(),
  offset = log(exposure),
  data   = long,
  method = "REML"
)

coef(fit_gam)[c("x1", "x2")]
summary(fit_gam)$edf   # effective df spent on the smooth baseline hazard
```

### Convergence of PEM toward Cox as $K \to \infty$

```{r convergence-check}
widths <- c(4, 2, 1, 0.5, 0.25, 0.1)

convergence <- do.call(rbind, lapply(widths, function(w) {
  cuts <- seq(w, 15, by = w)
  ll <- survSplit(
    Surv(time, status) ~ ., data = dat,
    cut = cuts, start = "tstart", end = "tstop",
    event = "event_split", episode = "interval"
  )
  ll$exposure <- ll$tstop - ll$tstart
  f <- glm(event_split ~ factor(interval) + x1 + x2,
           family = poisson(), offset = log(exposure), data = ll)
  b <- coef(f)[c("x1", "x2")]
  data.frame(
    interval_width = w,
    n_intervals     = length(cuts),
    x1_pem = b["x1"], x2_pem = b["x2"],
    x1_abs_diff = abs(b["x1"] - coef(fit_cox)["x1"]),
    x2_abs_diff = abs(b["x2"] - coef(fit_cox)["x2"])
  )
}))

convergence
```

**Expected pattern**: `x1_abs_diff` and `x2_abs_diff` should shrink
monotonically (not necessarily strictly) as `interval_width` decreases,
approaching zero as $K \to \infty$, per the profile-likelihood equivalence
derived above. The rate of convergence depends on how much within-interval
exposure heterogeneity there is at each width (the approximation
$o_{jk} \approx \Delta_k$ degrades for wide intervals with many
entries/exits/events mid-interval).

### Baseline cumulative hazard: three estimates, one true curve

```{r baseline-hazard-plot}
# True cumulative hazard from the simulation, at x = 0
H0_true <- function(t) (t / scale_true)^shape_true

# Cox: Breslow cumulative hazard (centered = FALSE -> at x = 0)
bh_cox <- basehaz(fit_cox, centered = FALSE)

# PEM: step-function cumulative hazard from the fitted interval intercepts.
# factor(interval) uses treatment contrasts, so level 1's absolute
# log-hazard is the intercept itself, and every other level's absolute
# log-hazard is intercept + its dummy coefficient.
cf         <- coef(fit_pem)
fac_names  <- grep("^factor\\(interval\\)", names(cf), value = TRUE)
alpha_hat  <- c(cf["(Intercept)"], cf["(Intercept)"] + cf[fac_names])
lambda_pem <- exp(as.numeric(alpha_hat))
widths_pem <- diff(c(0, cutpoints))
H0_pem     <- cumsum(lambda_pem * widths_pem)

# mgcv: integrate the fitted smooth hazard on a fine grid (x = 0).
# Setting exposure = 1 in newdata zeros the offset (log(1) = 0), so the
# "response"-scale prediction is the hazard rate itself, not a count.
tgrid   <- seq(0.01, 15, length.out = 500)
nd      <- data.frame(tmid = tgrid, x1 = 0, x2 = 0, exposure = 1)
haz_gam <- predict(fit_gam, newdata = nd, type = "response")
H0_gam  <- cumsum(haz_gam) * diff(tgrid)[1]

plot(bh_cox$time, bh_cox$hazard, type = "s", col = "black", lwd = 2,
     xlab = "Time", ylab = "Cumulative baseline hazard",
     main = "Three estimates of the baseline cumulative hazard")
lines(cutpoints, H0_pem, type = "s", col = "steelblue", lwd = 2)
lines(tgrid, H0_gam, col = "firebrick", lwd = 2)
curve(H0_true, add = TRUE, col = "grey40", lty = 2, lwd = 2)
legend("topleft",
       c("Cox (Breslow)", "PEM (step function)", "mgcv-Poisson (smooth)", "True"),
       col = c("black", "steelblue", "firebrick", "grey40"),
       lty = c(1, 1, 1, 2), lwd = 2)
```

**Expected pattern**: all three estimated curves should track the true
dashed curve closely, with the PEM curve visibly "steppy" at the chosen
interval width and the mgcv curve smooth by construction. The Cox
(Breslow) and PEM curves should nearly coincide, consistent with the
profile-likelihood equivalence above.

---

## Discussion

The equivalence derived here explains why PEM has historically been
attractive as "Cox regression using ordinary GLM software" (Whitehead
1980) — before efficient partial-likelihood implementations were common,
fitting a Poisson GLM on split data was a practical way to get
approximately the same $\hat\beta$. That motivation has faded now that
`coxph()` is fast and exact, but the person-interval data representation
it introduced turned out to be the right substrate for a different set of
extensions: once the baseline hazard is *any* parametric or penalized
function of $(t, x)$ rather than a free step function, the Poisson-GLM
machinery generalizes immediately to smooth time-varying effects, random
effects, and multi-state/competing-risks outcomes via standard GAM and
mixed-model software — which is precisely the PAMM research program (Bender
et al. 2018) and, in a different direction (smooth covariate effects but
exact Cox likelihood), `mgcv::gam(family = cox.ph())`.

For an NHANES-style application — smooth biomarker effects, a mortality
outcome, and finite sample sizes where an unpenalized step-function
baseline hazard would need many intervals to look smooth — the mgcv-Poisson
/ PAMM route offers a genuine alternative to `rms::cph()` + restricted
cubic splines for cases where a smooth *baseline hazard curve* itself
(not just smooth covariate effects) is of scientific interest, at the cost
of losing the exact semiparametric guarantee that makes Cox regression
attractive in the first place.

### Current limitations of this document

- The convergence and baseline-hazard chunks above have not been executed
  in the environment where this document was written (no R installation
  was available). The code has been checked carefully against `survival`
  and `mgcv` APIs, but treat the first run on your machine as a validation
  pass, not confirmation.
- The equivalence derivation assumes no ties within an interval beyond the
  single event used to define $R_k$; with `k`-tie intervals the argument
  generalizes by replacing the single-event score equation with a sum over
  the $k$ tied events, recovering the Poisson analogue of the
  Breslow-tie-corrected Cox likelihood.
- Survey-weighted extensions (relevant for NHANES analyses; see
  `vignettes/survey-weighted-survival.Rmd`) are not addressed here. The
  weighted Poisson-GLM literature (survey-weighted PEM) and a
  design-consistent version of PAMM are both open extensions parallel to
  the `svycph_fuse()` approach used elsewhere in this package.

---

## Session information

```{r session-info}
sessionInfo()
```
